---
title: "In-class template"
author: "Kimberly Wong"
format: pdf
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use and cleaning
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels
library(scales)

# read .xlsx files
library(readxl)

# visualize missing data
library(naniar)

# arrange plots
library(patchwork)

```


## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects"
)

# site information 
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality",
  na = c("", "n/a", "N/A", "over", "173+")
)

```

# 2. Cleaning 

## clean `sites` 

Goal: create a new object from `sites` that only includes NCOS sites

```{r}
#| label: ncos-sites

# create new object from sites
ncos_sites <- sites |> 
  # filter to only include NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")
```

## clean `taxon_list` 

Goal: convert character strings in the existing `taxon_list` data frame into snakecase

```{r}
#| label: cleaning-taxon

# create new object from taxon_list
taxon_list_clean <- taxon_list |> 
  # mutate is used to manipulate anything in a column
  # change column to all lowercase with underscores aka snakecase
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean `water_quality`

Goal: create a data frame with median pH, median DO (mg/L), median salinity (ppt) for each date and site 

- create a new object from `water_quality`
- clean column names using janitor
- only include surveys from quarterly NCOS sites
- create a new column called `date_site` to join with later data frames
- group by date_site
- calculate median pH, salinity, and dissolved oxygen
- ungroup the data frame
- filter out any outliers 

```{r}
#| label: cleaning-water-quality

# create a new object from water_quality
water_quality_clean <- water_quality |> 
  clean_names() |> 
  # filtering join to only include surveys from quarterly NCOS sites
  # only need to include right data frame since we are piping
  semi_join(ncos_sites, by = "site") |> 
  # create new column to join with later data frames
  unite("date_site", date, site, remove = TRUE) |> 
  # group by date_site
  group_by(date_site) |> 
  # calculate median pH, median salinity, median DO
  summarize(med_pH = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  # ungroup the data frame
  ungroup() |> 
  # filter our salinity outliers
  filter(date_site != "2022-08-12_NVBR")
  
```

## clean `aq_ins`

Goal: create a long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality information for each survey.

- create a new object from `aq_ins`
- clean column names
- filter to only include quarterly NCOS sites
- select columns
- filter to only include "filtered beaker" samples 
- convert the data frame to long format
- group by site, date, and taxon 
- calculate average abundance per liter (sum counts divided by 7.5 liters) 
- create a new column called `date_site` to join with water_quality
- join data frames
- join with taxon_list

```{r}
#| label: cleaning-aq-ins

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site, ostracod, copepod, hemiptera_corixidae_boatman, cladocera, nematode, diptera_ceratopogonidae, annelida_oligochaete, diptera_chironomid, annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer(
    # columns to make longer
    cols = ostracod:annelida_polychaete,
    # mew column for old columns name
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance"
  ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  ungroup() |> 
  unite("date_site", date, site, remove = FALSE) |> 
  left_join(water_quality_clean, by = "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel", 
    site == "NPB" ~ "South of Phelps Bridge", 
    site == "NPB1" ~ "North of Phelps Bridge"
    site == "NPB2" ~  "Phelps Road",
    site == "NWP" ~ "",
    site == "NDC" ~ "Devereux Creek"
  ))

```












